The Spatial and Temporal Domains of Modern Ecology

### 2017-06-22 Rmd/space-time.Rmd
Near final submission (JJC, PE, TT, JC edits)

# calr[DOI == "10.1007/s00442-005-0338-3" & feature == 2, 
#      mean(eff_dur, na.rm = T)]
calr <- calr[feature != 99, {
  lapply(.SD, function(x) mean(x, na.rm = TRUE))
},  by = .(DOI, feature), .SDcols = rnms[c(5:ncol(calr))]]
length(unique(calr$DOI)) + length(unique(dat$DOI))  # 134 papers providing dat
# unique(calr$DOI)

# calr_type <- data.table(cbind("DOI" = unique(calr$DOI), # calib field studies
#                               "study_type" = c(rep("field", 5), "paleo",
#                                                rep("field", 4))))

@@

# labels for temporal axes
# return interval
tlab1 <- c(expression(paste(NULL<="second"^-1)), expression("minute"^-1),
           expression("hour"^-1), expression("day"^-1), 
           expression("week"^-1), expression("month"^-1),            
           expression("year"^-1), expression("decade"^-1),
           expression("century"^-1), 
           expression("millenium"^-1), "unreplicated")# expression(infinity))
taxis1 <- cbind.data.frame( 
  c(1 / (24 * 60 * 60), 1 / (24 * 60), 1 / 24, 1, 7, 30, 365, 365 * 10, 
    365 * 100, 365 * 1000, 365 * 10000))
taxis1 <- data.table(taxis1)
setnames(taxis1, names(taxis1), "days")
taxis1[, logdays := log10(days)]

tlab2 <- c(expression(NULL<="second"), "minute", "hour", "day", "week", 
           "month","year","decade", "century", "millenium", "10 KA")  
taxis2 <- cbind.data.frame(c(1 / (24 * 60 * 60), 1 / (24 * 60), 1 / 24, 1, 7, 
                             30, 365, 365 * 10, 365 * 100, 365 * 1000, 
                             365 * 10000))
taxis2 <- data.table(taxis2)
setnames(taxis2, names(taxis2), c("days"))
taxis2[, logdays := log10(days)]

@@

datf[t_btwn_samp != 0, plot(log10(t_btwn_samp), log10(eff_ext))]
datf[, plot(log10(plot_res), log10(eff_dur))]
datf[, plot(log10(eff_dur), log10(eff_ext))]
datf[t_btwn_samp == 0, points(log10(eff_dur), log10(eff_ext), col = "red")]
datf[t_btwn_samp == 0 & log10(eff_dur) > 4]

>>>>

# calr[DOI == "10.1007/s00442-005-0338-3" & feature == 2, 
#      mean(eff_dur, na.rm = T)]
calr <- calr[feature != 99, {
  lapply(.SD, function(x) mean(x, na.rm = TRUE))
},  by = .(DOI, feature), .SDcols = rnms[c(5:ncol(calr))]]
length(unique(calr$DOI)) + length(unique(dat$DOI))  # 134 papers providing dat

@@

# labels for temporal axes
# return interval
tlab1 <- c("second", "minute", "hour", "day", "week", "month", "year", "decade",
           "century", "millenium", "unreplicated")
taxis1 <- cbind.data.frame( 
  c(1 / (24 * 60 * 60), 1 / (24 * 60), 1 / 24, 1, 7, 30, 365, 365 * 10, 
    365 * 100, 365 * 1000, 365 * 10000))
taxis1 <- data.table(taxis1)
setnames(taxis1, names(taxis1), "days")
taxis1[, logdays := log10(days)]

tlab2 <- c("second", "minute", "hour", "day", "week", 
           "month","year","decade", "century", "millennium", "10 KA")  
taxis2 <- cbind.data.frame(c(1 / (24 * 60 * 60), 1 / (24 * 60), 1 / 24, 1, 7, 
                             30, 365, 365 * 10, 365 * 100, 365 * 1000, 
                             365 * 10000))
taxis2 <- data.table(taxis2)
setnames(taxis2, names(taxis2), c("days"))
taxis2[, logdays := log10(days)]

@@

```
<a href="#top">Back to top</a>

```{r, eval=FALSE, echo=FALSE}
# check effect of averaging by DOI
datf[, {
 list("a" = log10(mean(plot_res)), "b" = log10(mean(eff_ext, na.rm = TRUE)))
}, by = DOI][, list("a" = mean(a), "b" = mean(b, na.rm = TRUE))]
datf[, list("a" = mean(log10(plot_res)), "b" = mean(log10(eff_ext), na.rm = T))]

### 2017-06-22 paper/SI.tex
Near final submission (JJC, PE, TT, JC edits)

\section*{Supplementary Information}% Make the title.
\subsection*{Methods}% Make the title.
\vspace{-4pt}
%\maketitle 
\subsubsection*{Paper selection}
\vspace{-10pt}
We used the 2012 Web of Science impact factors to select the highest ranked ecology-themed journals that published studies with an observational component, excluding journals devoted to reviews, meta-analyses, laboratory, cellular, or experimental studies.  To select a representative sample of recent ecology studies, we downloaded the metadata for all papers published in the selected journals (Table S1) between 2004 and 2014. Our study involved six different observers (those reviewing the papers to extract the observational scales), each of whom was given a randomly selected batch of 500 titles.  A separate set of 20 papers was also randomly selected, and this set was given to all observers to review, in order to 1) calibrate the interpretations and extraction of scale-related information between observers, and 2) to estimate between-observer variance.

@@

  \Hitem \emph{\textbf{What is sampling duration, and how do we record it?}} How long an individual observation of an individual point in space took to make. Sampling duration multiplied by the number of repeats observations is used to calculate study duration. Often this value will not be reported, so you will have to use your best judgement, based on your knowledge of ecological methods, to approximate the duration. For example, for a field based method with intensive plot methods, if you can't estimate a plausible duration, assign a token 1 day. For remote sensing observations you can assume one second (the observations are effectively instantaneous).  
   \Hitem \emph{\textbf{What is the study duration, and how do estimate it?}} The study duration (or, simply, duration) is the total period of time over which the phenomenon of interest was observed. More specifically, in the case of repeated observations, this is the total time that elapsed between the first and last observations at a given point in space (or of the same individual organism or community) were made. For once-off (unreplicated) observations, this time is equivalent to the sample duration. However, there may be cases where once-off observations may have a longer duration then the sample duration.  For example, consider a study that counts occurrences of pollinators over three years, using transects that are located in different locations within the broader study area during each year \cite{rollin_differences_2013-1}. The observations are therefore not strictly temporally replicated, but the authors control for year of collection in their subsequent analysis to avoid confounding effects.  In this case, we can consider the effective duration to be three years, as the temporal information is encoded in the analysis. 

@@

    \Hitem \emph{\textbf{What about intervals and durations for instrument-collected, or automatically-collected, observations?}} These are often similar in essence to the paleo-reconstruction case. Take the minimum temperature or daily rainfall recorded at a weather station, which require constant observation over 24 hours to report. In such cases, the interval and sampling duration are both 24 hours. On the other hand, automated logging systems will provide a series of high frequency observations that are collected instantaneously. In these cases the sampling duration should be $\leq$second, and the interval should be the period between successive instantaneous measurements. 

@@

  \Hitem \emph{\textbf{What is study extent, and how do I record it?}} Study extent (or, simply, extent) is defined as the total extent bounding all spatial replicates, divided by 10,000 to convert to hectares. For studies in which spatial replicates are not spatially contiguous, this means the area of the minimum polygon bounding all spatial replicates. For example, if the study is conducted in a national park, the effective survey extent would be the area of the national park; if the study is conducted in three national parks, the effective survey extent would be the sum of the areas of all national parks. To calculate the effective survey extent, use the area of the study area/region given in the paper; when the area is not given, but when the survey region is given by name (e.g., Joshua Tree National Park or United States), look up the area through an online search and convert to hectares. When the area is not named, but a map is given, use an appropriate digitizing platform with a suitable map-providing backend (e.g. Google Earth Pro, QGIS with OpenLayers plugin) to navigate to the region and delineate a minimum convex polygon surrounding the plots/transects/sampling units to calculate the area in hectares. When studies focus on portions of survey regions that are clearly distinct from their surroundings (e.g., mangroves in a coastal National Park), try to delineate the focal portions (the mangroves) and not the larger survey region (entire National Park) using Google Earth Pro or a similar application with digitizing and area estimation capabilities.

For spatially contiguous studies (e.g. those based on satellite imagery), the extent is the total area covered by the imagery (in such cases, extent equals actual extent), but only records the area of imagery analyzed by the authors (e.g. if the study area required four Landsat scenes to cover, but covered only the inner quarter of each image, report the extent as the summed area of the four quarters). However, if spatially contiguous studies only use a sub-sample of pixels, extent is the area of the minimum convex polygon bounding all pixels (calculated following the methods above).

@@

  \Hitem \emph{\textbf{How do you determine sampling resolution for paleo-reconstructions and other approaches where a sampling method is presumed to draw from a larger area (e.g. mammal traps, mist nets, etc)?}}  For sampling resolution, estimate the size of the sample actually taken, rather than the assumed catchment/shed area of the sample (e.g. the area of the corer used to take a sediment sample, rather than an estimate of the area that that sample is assumed to draw from), and then indicate that the plot resolution was uncertain. Related to this, you may also have to estimate the number of samples collected, as exemplified in a charcoal study of Europe where the number of lakes sampled was provided, not the number of cores per lake \cite{molinari_exploring_2013}.

@@

\subsubsection*{Accounting for estimation uncertainty}
\vspace{-10pt}
Two major sources of uncertainty affected our estimation of observational scales: 1) unclear documentation of observational scales in the reviewed studies, 2) variation between observers in estimating observational scales. We estimated and accounted for these uncertainties in several ways. First, we calculated the degree of between-observer variability based on each observer's results from reviewing the 20-paper calibration set. We calculated how well observers agreed regarding paper inclusion/exclusion, how many extractable observations there were per included paper, and what the coefficient of variation was across all observers' estimates of scale. We also recorded when observations were reported in any study (calibration set or otherwise) with an unclear or missing scale value.

We used the between-observer coefficients of variation for each dimension as the basis for randomly perturbing the scale values for each of the 371 observations over 1000 iterations. For each observation at each iteration, we perturbed its scale value in each dimension by randomly selecting (from a uniform distribution) a percentage value p that fell between 100 + y and 100 - y, where y was the between-observer coefficient of variation (expressed as a percentage) for a given observational dimension, and multiplying that observation's scale value by that proportion. This perturbed set of observations provided a basis for estimating uncertainty within our extracted set of observational scales.

@@

\subsubsection*{Extracting and analyzing data from earlier meta-analyses}

419 observational studies lasting $\leq100 years; longer studies were excluded because the upper duration bound was not published)

reviewed in this study

@@

Second, we calculated the intra-class correlation coefficient \cite{bartko_intraclass_1966} to assess agreement between observers regarding the number of ecological observations that could be extracted from each paper (multiple ecological observations were reported in many studies, and we listed observations as separate records if they varied on one or more dimensions).  The coefficient, calculated using the IRR package \cite{gamer_irr:_2012} of R, was 0.71 (F = 15.4, p$<$0.0001, 95\% confidence interval = 0.54 - 0.85)

@@

\begin{figure}[ht]
%\begin{wrapfigure}{c}{1\textwidth}
\center{\includegraphics[width=0.8\textwidth]{figures/dim_by_yr.png}}
\vspace{5 pt}
\caption{Trends in observational scales by year of publication. The coefficient of a weighted (by number of studies in each year) linear regression, fit to the logarithm (base 10) of the mean scale values (calculated for each publication year) for the six assessed dimensions ispresented at the top of each plot, as well as the model coefficient of determination and p-value. }
\label{afoto1}
\end{figure}
%\clearpage

\subsubsection*{Choice of bandwidth in kernel density estimation}
\vspace{-10pt}
Figure S3 indicates the effect that varying bandwidth has on the appearance of the kernel density estimates. The smallest bandwidth (0.4) tested (Fig. S3 top) shows that the primary observational concentrations described in the main text are evident, but tend to be devolved into separate concentrations. An example is the oblong concentration of observations that in Fig. 1A (and in the lower left panel in Fig. S3) is bounded on the lower right at monthly to yearly intervals and 100-1000 m$^2$ resolutions and on the upper left by near-daily to monthly observations and 0.1-10 ha resolutions. With the smaller bandwidth this concentration appears as two separate patches (Fig. S2 top left), but with 0.7 bandwidth applied becomes coherent (Fig. S3 middle left).  This increasing separation of points into separate clusters is expected as bandwidth is reduced.

>>>>

\section*{Supplementary Information}% Make the title.

The following text provides additional information on this study's Methods and Results.  The full manuscript, data, and code used to analyze the data are available on an accompanying github repository (https://github.com/ecohydro/ecoscales). \textbf{Note to editors and reviewers: this repository is currently private, but will be transitioned to a public repository upon acceptance of manuscript.}

\subsection*{Methods}% Make the title.
\vspace{-4pt}
%\maketitle 
\subsubsection*{Paper selection}
\vspace{-10pt}
We used the 2012 Web of Science impact factors to select the highest ranked ecology-themed journals that published studies with an observational component, excluding journals devoted to reviews, meta-analyses, laboratory, cellular, or experimental studies.  To select a representative sample of recent ecology studies, we downloaded the metadata for all papers published in the selected journals (Table S1) between 2004 and 2014. Our study involved six different observers (those reviewing the papers to extract the observational scales), each of whom was given a randomly selected batch of 500 titles.  A separate set of 20 papers was also randomly selected, and this set was given to all observers to review independently, in order to 1) calibrate the interpretations and extraction of scale-related information between observers, and 2) to estimate between-observer variance.

@@

  \Hitem \emph{\textbf{What is sampling duration, and how do we record it?}} How long an individual observation of an individual point in space took to make. Sampling duration multiplied by the number of repeats observations is used to calculate study duration. Often this value will not be reported, so you will have to use your best judgement, based on your knowledge of ecological methods, to approximate the sampling duration. For example, for a field based method with intensive plot methods, if you can't estimate a plausible sampling duration, assign a token 1 day. For remote sensing observations you can assume one second (although the observations are effectively instantaneous).  
   \Hitem \emph{\textbf{What is duration, and how do we estimate it?}} The duration is the total period of time over which the phenomenon of interest was observed. More specifically, in the case of repeated observations, this is the total time that elapsed between the first and last observations at a given point in space (or of the same individual organism or community). For once-off (unreplicated) observations, this time is equivalent to the sample duration. However, there may be cases where once-off observations may have a longer duration than the sample duration.  For example, consider a study that counts occurrences of pollinators over three years, using transects that are located in different locations within the broader study area during each year \cite{rollin_differences_2013-1}. The observations are therefore not strictly temporally replicated, but the authors control for year of collection in their subsequent analysis to avoid confounding effects.  In this case, we can consider the effective duration to be three years, as the temporal information is encoded in the analysis. 

@@

    \Hitem \emph{\textbf{What about intervals and durations for instrument-collected, or automatically-collected, observations?}} These are often similar in essence to the paleo-reconstruction case. Take the minimum temperature or daily rainfall recorded at a weather station, which require constant observation over 24 hours to report. In such cases, the interval and sampling duration are both 24 hours. On the other hand, automated logging systems will provide a series of high frequency observations that are collected instantaneously. In these cases the sampling duration should be a token one second (to keep consistent with remote sensing [Q7]), and the interval should be the period between successive instantaneous measurements. 

@@

  \Hitem \emph{\textbf{What is extent, and how do I record it?}} Extent is defined as the total area enclosed within a perimeter defined by the outermost spatial replicates, divided by 10,000 to convert to hectares. For studies in which spatial replicates are not spatially contiguous, this means the area of the minimum polygon bounding all spatial replicates. To calculate the effective survey extent, use the area of the study area/region given in the paper; when the area is not given, but when the survey region is given by name (e.g., Joshua Tree National Park or United States), look up the area through an online search and convert to hectares, provided the observations were collected throughout the named survey region. When the area is not named, but a map is given (or if the area is named and the map of replicates shows that the replicates were located in a smaller area of the named region), use an appropriate digitizing platform with a suitable map-providing backend (e.g. Google Earth Pro, QGIS with OpenLayers plugin) to navigate to the region and delineate a minimum convex polygon surrounding the replicates (plots/transects/other sampling units) to calculate the area in hectares. There is an exception to this rule, however, for studies that focus on features that are clearly distinct and functionally isolated from their surroundings. Examples of such cases might be mangrove forests patches within in a coastal National Park, or populations of a rare species confined to three disjunct protected areas. In these cases, the extent will be the summed area of the sampled units (the summed area of sampled mangrove stands, and the summed areas of the three protected areas). In other words, try to delineate the focal portions and not the larger survey region.

For spatially contiguous studies (e.g. those based on satellite imagery), the extent is the total area covered by the imagery (in such cases, extent equals actual extent), but only record the area of imagery analyzed by the authors (e.g. if the study area required four Landsat scenes to cover, but covered only the inner quarter of each image, report the extent as the summed area of the four quarters). However, if spatially contiguous studies only use a sub-sample of pixels, extent is the area of the polygon enclosing the outermost pixels (calculated following the methods above).

@@

  \Hitem \emph{\textbf{How do you determine resolution for paleo-reconstructions and other approaches where a sampling method is presumed to draw from a larger area (e.g. mammal traps, mist nets, etc)?}}  For sampling resolution, estimate the size of the sample actually taken, rather than the assumed catchment/shed area of the sample (e.g. the area of the corer used to take a sediment sample, rather than an estimate of the area that that sample is assumed to draw from), and then indicate that the plot resolution was uncertain. Related to this, you may also have to estimate the number of samples collected, as exemplified in a charcoal study of Europe where the number of lakes sampled was provided, not the number of cores per lake \cite{molinari_exploring_2013}.

@@

\subsubsection*{Accounting for estimation uncertainty}
\vspace{-10pt}
Two major and related sources of uncertainty affected our estimation of observational scales: 1) unclear documentation of observational scales in the reviewed studies, 2) variation between observers in estimating observational scales (largely in cases where scales were not explicitly reported). We estimated and accounted for these uncertainties in several ways. First, we calculated the degree of between-observer variability based on each observer's results from reviewing the 20-paper calibration set. We calculated how well observers agreed regarding paper inclusion/exclusion, how many extractable observations there were per included paper, and what the coefficient of variation was across all observers' estimates of scale. We also recorded when observations were reported in any study (calibration set or otherwise) with an unclear or missing scale value.

We used the between-observer coefficients of variation for each dimension as the basis for randomly perturbing the scale values for each of the 371 observations over 1000 iterations. For each observation at each iteration, we perturbed its scale value in each dimension by randomly selecting (from a uniform distribution) a percentage value p that fell between 100 + y and 100 - y, where y was the between-observer coefficient of variation (expressed as a percentage) for a given observational dimension, and multiplying that observation's scale value by that proportion. In certain cases, the perturbation resulted in values for some certain dimensions that were not physically possible (e.g. interval longer or actual duration longer than duration; actual extent larger than extent), in which case we adjusted the perturbed values for the dimension that should be the smallest to equal that of the one that should be largest (e.g. we set interval to equal the perturbed duration value). This perturbed set of observations provided a basis for estimating uncertainty within our extracted set of observational scales.

@@

\subsubsection*{Extracting and analyzing data from earlier meta-analyses}
To compare the results of our analysis with the observational scales of earlier ecological studies, we used graph capture software\footnote{http://arohatgi.info/WebPlotDigitizer/app/} to extract the data values from Figure 6.1 in Tilman \cite{tilman_ecological_1989}, Figure 1 in Kareiva and Anderson \cite{kareiva_spatial_1988}, and Figure 2 in Porter et al \cite{porter_crop_2005}. To maintain as much comparability as possible with our inclusion criteria, we excluded experiment studies in Tilman's data, as well as the values of any studies having duration values greater than 100 years (no upper time bound was provided for these), leaving duration values for 419 (out of 623) studies. As Tilman presented duration values as a histogram, we calculated the mean duration across all studies as the weighted (by number of observations per bin) mean of bin center-point values (i.e. the weighted mean of the bin means).  We also excluded four (of 29) observation values from Porter et al's data on observational extent and frequency, which, in contrast to the other 25, were not randomly selected. Porter et al also used irregular scales for both their X (frequency) and Y (extent) axis, therefore we had to visually estimate the scale values for each data point after graphical extraction, and converted their extent values (in km) to hectares and their frequency values to intervals. Kareiva and Anderson presented resolution as plot diameters (m), which we squared to make comparable to our resolution metric.

@@

Second, we calculated the intra-class correlation coefficient \cite{bartko_intraclass_1966} to assess agreement between observers regarding the number of ecological observations that could be extracted from each paper (multiple ecological observations were reported in many studies; we listed observations as separate records if they measured different features, or more rarely, varied substantially on one or more dimensions).  The coefficient, calculated using the IRR package \cite{gamer_irr:_2012} of R, was 0.71 (F = 15.4, p$<$0.0001, 95\% confidence interval = 0.54 - 0.85)

@@

\begin{figure}[ht]
%\begin{wrapfigure}{c}{1\textwidth}
\center{\includegraphics[width=0.8\textwidth]{figures/dim_by_yr.png}}
\vspace{5 pt}
\caption{Trends in observational scales by year of publication. The coefficient of a weighted (by number of studies in each year) linear regression, fit to the logarithm (base 10) of the mean scale values (calculated for each publication year) for the six assessed dimensions is presented at the top of each plot, as well as the model coefficient of determination and p-value. }
\label{afoto1}
\end{figure}
%\clearpage

\subsubsection*{Choice of bandwidth in kernel density estimation}
\vspace{-10pt}
Figure S3 indicates the effect that varying bandwidth has on the appearance of the kernel density estimates. The smallest bandwidth (0.4) tested (Fig. S3 top) shows that the primary observational concentrations described in the main text are evident, but tend to be devolved into separate concentrations. An example is the oblong concentration of observations that in Fig. 1A (and in the lower left panel in Fig. S3) is bounded on the lower right at monthly to yearly intervals and 100-1000 m$^2$ resolutions and on the upper left by near-daily to monthly observations and 0.1-10 ha resolutions. With the smaller bandwidth this concentration appears as two separate patches (Fig. S3 top left), but with 0.7 bandwidth applied becomes coherent (Fig. S3 middle left).  This increasing separation of points into separate clusters is expected as bandwidth is reduced.  However, this assessment indicates that the choice of bandwidth does not alter the interpretation of the results, as the domains of highest and lowest concentrations remain the same.
